Learning Theory Notes

### 2022-05-30 notebooks/theory_notes.tex
changes to notes

Our goal is to approximate the function $B = f(X)$ with some learned function
$\widehat{B} = \hat{f}(X)$ from evolutionary simulations. In each evolutionary
simulation, we sample some evolutionary parameters $X_k$ and evolve $r$
independent populations forward in time under these parameters and observe some
genealogy. From this genealogy, we estimate the reduction in neutral diversity
as:

\begin{align}
  \bar{B}_k = \frac{1}{r} \sum_{i=1}^r \frac{\hat{\pi}_{k,i}}{4N\mu}
\end{align}

there $\hat{\pi}$ is Tajima's estimator for pairwise diversity within a tree.
When we estimate $\hat{\pi}$ from using branch-mode tree statistics from an
observed genealogy, $\mu \to 1$, so we can ignore mutation.  If we take
expectation over the evolutionary process,

\begin{align}
  \E(\bar{B}_k) &= \frac{1}{4Nr} \sum_{i=1}^r \E(\hat{\pi}_{k,i}) \\
                &= \frac{T_k^{(2)}}{4N}  \\
                &= \frac{4B_kN}{4N}  \\
                &= B_k  \\
\end{align}

thus, the estimated reduction in diversity from simulations $\bar{B}$ is
unbiased, since $\E(\hat{\pi}_k) = 2T_k^{(2)} = 4B_k N$. For independent
evolutionary replicates, we calculate the variance of this estimator using
Tajima's equation for the variance of $\hat{\pi}$ as

\begin{align}
  \var(\bar{B}_k) &= \frac{1}{16 N^2 r^2} \sum_{i=1}^r \var(\hat{\pi}_{k,i}) \\
                  &= \frac{\var(\hat{\pi}_{k,i})}{16 N^2 r} \\
                  &= \frac{1}{16N^2 r} \left( \frac{n + 1}{3(n-1)}\theta  + \frac{2(n^2 + n + 3)}{9n(n-1)}\theta^2 \right) \\
                  &= \frac{1}{16N^2 r} \left( \frac{n + 1}{3(n-1)} 4BN + \frac{2(n^2 + n + 3)}{9n(n-1)} 16B^2N^2 \right) \\
                  &= \underbrace{\frac{n + 1}{12(n-1)} \frac{B}{N r}}_\text{sampling noise} + \underbrace{\frac{2(n^2 + n + 3)}{9n(n-1)} \frac{B^2}{r}}_\text{evolutionary variance}
\end{align}

@@

\begin{align}
  \lim_{N \to \infty} \P\left(|\bar{B}_N - B| \ge \epsilon \right)  &\le \lim_{N \to \infty} \frac{\var(\bar{B}_N)}{\epsilon^2} \\
  \lim_{N \to \infty} \P\left(|\bar{B}_N - B| \ge \epsilon \right)  &\le \frac{2B^2}{9 r \epsilon^2}. \\
\end{align}

Thus, even if we sample the entire population, and let the population size $N
\to \infty$, $\bar{B}_n$ is still an inconsistent estimator in $n$. Intuitively
this is because

In our case, we estimate pairwise diversity from the genealogical tree of the
entire population, so $n = 2N$

\begin{align}
  \var(\bar{B}_k) &= \frac{n + 1}{12(n-1)} \frac{B}{N r} + \frac{2(n^2 + n + 3)}{9n(n-1)} \frac{B^2}{r} \\
                  &= \frac{2N + 1}{12(2N-1)} \frac{B}{N r} + \frac{2(4N^2 + 2N + 3)}{18N(2N-1)} \frac{B^2}{r} \\
                  &\approx \frac{B}{12N r} + \frac{2N^2 + N}{9N^2} \frac{B^2}{r}
\end{align}

@@

\begin{align}
  \lim_{N \to \infty} \var(\bar{B}_k) &= \frac{2B^2}{9r}
\end{align}

>>>>

Our goal is to approximate a function that describes the reduction in pairwise
diversity (relative to the neutral expectation) $B$ due to BGS for some set of
evolutionary parameters $\mathbf{x}$. The true reduction function is a
conditional expectation $B(\mathbf{x}) = \E[T_2(\mathbf{x}) | \mathbf{x}]$
taken over an infinite number of evolutionary replicates. Here
$T_2(\mathbf{x})$ is the pairwise coalescent rate under BGS for parameters
$\mathbf{x}$, with time scaled in $2N$ generation units.

We approximate this function $B(\mathbf{x})$ with $\widehat{B}(\mathbf{x})$
from evolutionary simulations using statistical learning. In each evolutionary
simulation, we sample some parameter set $\mathbf{x}$ and evolve $r$
independent populations forward in time and observe the resulting genealogy at
a neutral site a given recombination distance away (this is one of the
parameters). From this full population genealogy of $2N$ gametes, we estimate
the reduction in neutral diversity as

\begin{align}
  \bar{B} = \frac{1}{r} \sum_{i=1}^r \frac{\hat{\pi}_{i}}{4N\mu}
\end{align}
%
where $\hat{\pi}$ is Tajima's estimator for pairwise diversity within a tree.
For now, we can imagine letting the neutral mutations saturate as $\mu \to 1$
so we can ignore mutation rate hereafter. If we take expectation over the evolutionary
process,

\begin{align}
  \E[\bar{B}] &= \frac{1}{4Nr} \sum_{i=1}^r \E[\hat{\pi}_{i}] \\
                          &= \frac{2T_2}{4N}  \\
                          %&= \frac{4B(\mathbf{x})N}{4N}  \\
                          &= B
\end{align}
%
thus, the estimated reduction in diversity from simulations $\bar{B}$ is
unbiased, since $\E[\hat{\pi}] = 2T_2 = 4BN$.

\subsection*{The variance of the estimator $\bar{B}$}

Using the law of total variance, we can write the variance of $\bar{B}$ as

\begin{align}
  \var(\bar{B}) = \underbrace{\E(\var(\bar{B} | \mathcal{G}))}_\text{sampling noise} + \underbrace{\var(\E(\bar{B} | \mathcal{G}))}_\text{evolutionary variance}
\end{align}

where $\mathcal{G}$ is the genealogy at the focal neutral site affected by BGS
at some segment. The distribution of genealogies $P(\mathcal{G})$ is unknown
and quite complex under BGS, but we can use a simplifying assumption to get a
rough estimate of the variance of $\bar{B}$.

Under strong BGS (e.g. $\nicefrac{\mu}{s}N \gg 1$), the genealogy is often
modeled as a neutral coalescent process with rescaled with an effective
population size $N_e = BN$ (though see Cvijovi\'{c} et al. 2018). In the weak
selection limit $s \to 0$, approximation becomes exact as BGS becomes
effectively neutral. In both regimes, selection is treated as rescaling a
neutral genealogical process to $N_e = BN$. We refer to this the
\textbf{neutral-BGS model}, and it allows for us to approximate $\var(\bar{B})$
with some $\var(\bar{B}')$ where $\bar{B}'$ is the mean reduction if we
averaged over neutral genealogies from a population of $BN$
diploids, rather than averaged over exact selection simulations. Note that at
$B=1$, the neutral-BGS model becomes exact, as there is no reduction in
diversity due to background selection and the only evolutionary variance is
through the neutral coalescent process.

If the genealogies are neutral, we can calculate the variance of $\bar{B}$
using Tajima's equation for the variance of $\hat{\pi}$. For $n$ samples and
$r$ evolutionary replicates,

\begin{align}
  \var(\bar{B}') &= \frac{1}{16 N^2 r^2} \sum_{i=1}^r \var(\hat{\pi}_{i}) \\
                &= \frac{\var(\hat{\pi}_{i})}{16 N^2 r} \\
                  &= \frac{1}{16N^2 r} \left( \frac{n + 1}{3(n-1)}\theta  + \frac{2(n^2 + n + 3)}{9n(n-1)}\theta^2 \right) \\
                  &= \frac{1}{16N^2 r} \left( \frac{n + 1}{3(n-1)} 4BN + \frac{2(n^2 + n + 3)}{9n(n-1)} 16B^2N^2 \right) \\
                  &= \underbrace{\frac{n + 1}{12(n-1)} \frac{B}{N r}}_\text{sampling noise} + \underbrace{\frac{2(n^2 + n + 3)}{9n(n-1)} \frac{B^2}{r}}_\text{evolutionary variance}
\end{align}

@@

\begin{align}
  \lim_{N \to \infty} \P\left(|\bar{B}_N - B| \ge \epsilon \right)  &\le \lim_{N \to \infty} \frac{\var(\bar{B}_N)}{\epsilon^2} \\
  \lim_{N \to \infty} \P\left(|\bar{B}_N - B| \ge \epsilon \right)  &\le \frac{2B^2}{9 r \epsilon^2}. \label{eq:bound}
\end{align}

Thus, even if we sample the entire population, and let the population size $N
\to \infty$, $\bar{B}_n$ is still an inconsistent estimator in $n$. Intuitively
this is because the evolutionary variance is not decreased by taking more
samples, \emph{only by taking more evolutionary replicates}.

Below I work through where this bound came from. If we estimate pairwise
diversity from the genealogical tree of the entire population such that $n =
2N$,

\begin{align}
  \var(\bar{B}) &= \frac{n + 1}{12(n-1)} \frac{B}{N r} + \frac{2(n^2 + n + 3)}{9n(n-1)} \frac{B^2}{r} \\
                &= \frac{2N + 1}{12(2N-1)} \frac{B}{N r} + \frac{2(4N^2 + 2N + 3)}{18N(2N-1)} \frac{B^2}{r} \\
                &\approx \frac{B}{12N r} + \frac{2N^2 + N}{9N^2} \frac{B^2}{r}
\end{align}

@@

\begin{align}
  \lim_{N \to \infty} \var(\bar{B}) &= \frac{2B^2}{9r}
\end{align}

In practice, we use a different, better estimator of pairwise diversity than
Tajima's $\pi$: branch statistic pairwise diversity (Ralph 2019, Ralph et al.
2020). This estimator removes mutation as a source of sampling noise and since
$n = 2N$, our only source of variance in $\bar{B}$ is in the evolutionary
process.

% Note that since the bound in Equation \eqref{eq:bound} is a probability, it
% must be that $\frac{2B^2}{9 r \epsilon^2} \le 1$. Thus this equation is limited
% to considering bounds $\epsilon$ that are

% \begin{align}
%   \sqrt{\frac{2}{9r}} B \le \epsilon \\
% \end{align}

% For $B \approx 1$, $\sqrt{\frac{2}{9r}} \le \epsilon$, such that if $r = 50$,
% we can only consider bounds $\epsilon \ge \sqrt{\nicefrac{1}{225}} \approx
% 0.067$ because Chebyshev isn't a particularly sharp inequality.

\subsection*{Partitioning the Prediction Error}

To learn the function $B(\mathbf{x})$, we run evolutionary simulations under a
training set of parameters, giving us $\{(\mathbf{x}_1, \bar{B}_1),
(\mathbf{x}_2, \bar{B}_2), \ldots, (\mathbf{x}_n, \bar{B}_n)\}$, where each
$\bar{B}_i$ is the observed reduction over $r$ replicates. Since our goal is to
approximate $\hat{B}(\mathbf{x})$ as closely as possible to $B(\mathbf{x})$
over the distribution of $\mathbf{x}$ most close to that across the genome, we
find some $\hat{B}(\mathbf{x})$ that minimizes expected loss. An important
metric, even if training is conducted with a different loss, is the mean
squared error

\begin{align}
  MSE(\bar{B}, \hat{B}) = \E[(\bar{B}(\mathbf{x}) - \hat{B}(\mathbf{x}))^2].
\end{align}

We can think of $\bar{B}(\mathbf{x})$ as a true reduction $B(\mathbf{x})$ plus
a deviation $\varepsilon$ due to the particular evolutionary replicates
simulated, e.g. $\bar{B}(\mathbf{x}) = B(\mathbf{x}) + \varepsilon$.

\begin{align}
  MSE(\bar{B}, \hat{B}) &= \E[(\bar{B}(\mathbf{x}) - \hat{B}(\mathbf{x}))^2] \\
                        &= \E[(B(\mathbf{x}) + \varepsilon - \hat{B}(\mathbf{x}))^2] \\
                        &= \underbrace{\var(\hat{B}(\mathbf{x}))}_\text{training variance} + 
                      \underbrace{\var(\varepsilon)}_\text{evolutionary variance} + \underbrace{\E[\hat{B}(\mathbf{x}) - B(\mathbf{x})]^2}_{\mathrm{bias}^2}
\end{align}

At $B=1$, the evolutionary variance is exactly the neutral variance given by
$\var(\varepsilon) = \nicefrac{2B^2}{9r}$. Thus, MSE loss has an absolute bound at
$B=1$,

\begin{align}
  MSE(\bar{B}, \hat{B} | B = 1) \ge \frac{2}{9r}
\end{align}

For $B < 1$, we can write the evolutionary variance as some deviation
away from the coalescent noise under the neutral-BGS model $\var(\varepsilon) =
\var(\bar{B}') + e$.

% For a particular training set
% $\mathcal{T} = \{\mathbf{X}_\text{train}, \mathbf{B}_\text{train}\}$, the mean
% squared error is
